\textls[-0]{\thetitle}

### 2023-03-07 canonicitymodel.tex
Working on appendices.

Now we prove the consistency and canonicity using canoncity model. We follow the reducibility argument from~\cite{kaposi2019gluing, coquand2018canonicity,sterling2019algebraic} to construct the canonicity model.
Each of their proof is based on QIIT, Category with Family\cite{dybjer1995internal} and Generalized Algebraic Theory \cite{cartmell1986generalised} as the syntax. However, our proof will still work directly on the earlier introdcued syntax for readability.
We will first specify the formulation of the most basic canonicity theorem.

@@

Canonicity is one of the basic criteria to consider a dependent type theory to be ``a programming language or as a computational foundation for mathematics''.\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
We can even argue that, if this theorem is proven in a computable meta-logic, 
then by Curry-Howard Correspondence, this theorem provides a big-step interpreter for closed term of boolean type.\YZ{Can we say something about strong/weak normalization?}\EDJreply{
  No. \\
  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
  4. Of course we can conjecture we have that normalization function \\
  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
  but we don't have termination (a runnable interpreter) for open terms.  \\ 
  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
}

>>>>

Now we prove the consistency and canonicity using canoncity model. We follow the reducibility argument from~\cite{kaposi2019gluing, coquand2018canonicity,sterling2019algebraic} to construct the canonicity model.
Each of their proof is based on QIIT, Category with Family\cite{dybjer1995internal} and Generalized Algebraic Theory \cite{cartmell1986generalised} as the syntax. Meanwhile, our proof will still work directly on the earlier introdcued syntax for readability.
We will first specify the formulation of the most basic canonicity theorem.

@@

Canonicity is a basic criteria to consider a dependent type theory to be ``a programming language or as a computational foundation for mathematics''.\footnote{See nlab explanation: \href{https://ncatlab.org/nlab/show/canonical+form}{Canonical Form}}
We can even argue that, if this theorem is proven in a computable meta-logic, 
then by Curry-Howard Correspondence, this theorem provides a big-step interpreter for closed term of boolean type.\YZ{Can we say something about strong/weak normalization?}\EDJreply{
  No. \\
  1. In the declarative system, we cannot talk about strong/weak normalization (because we don't have operational semantic, if you recall the definition of strong/weak normalization, it based on reduction/path of steps/op semantic). So any review says strong/weak normalization to you, please stop them. \\ 
  2. In declarative system, we only say ``normalization'' directly, which means, we construct a function that \\
      for arbitrary class of (judgemental-equivalent) terms, it will return one syntax representation (called normal-form) for each class \\
    (apparently most people will make the normal form  without beta redex, however, interestingly, they may do eta expansion instead of eta reduction because theoretically easier) \\ 
  3. So translation won't work because it compiles away the normal form of linkage (the $\mu^+$). This part requires a direct proof which construct normal form and neutral form for our calculus, and a normalization model from which we extract a runnable interpreter for open terms \\
  4. Of course we can conjecture we have that normalization function \\
  5. Also we do have close term normalization, which is canonicity. Or more operationally speaking, we have termination (a runnable interpreter) for close term (since QIIT is runnable, the canoncity model itself is the runnable interpreter), \\
  but we don't have termination (a runnable interpreter) for open terms.  \\ 
  So earlier you say ``normalize'' a recursion on an inductive type, that is no doable -- if the programmer wants to program on meta-theory, they will have to manually write out the a chain of judgemental equality (with a bunch of beta rules inside) by themselves. 
}

### 2023-03-07 syn-translate.tex
Working on appendices.

The implementation described in \cref{sec:coqimpl} has hinted the fact that
linkage can be translated into dependent tuple type.
The described compilation strategy can be considered as
a \textit{type preserving syntactic translation}, a function mapping from our syntactic
model to a ``subpart''\footnote{Not exactly subpart---the only
difference lies in "Wrec". In our metatheory, it is constructed
using linkage; in ``subpart'', it will be using $\Sigma$-types (translation
of the linkage).} of our syntactic model \textit{without
linkages}.

@@

The main idea of this syntactic translation is that, we make sure
$\goodSig{\Gamma}{\nu^+ \_ }{n+1}$ and $\goodSig{\Gamma}{\mu^+ \_}{n+1}$
wrapped into a dependent pair. These two definitions decide the
syntactic translation for $\goodSig{\Gamma}{\_}{n}$ and $\cL$. The
syntactic translation for $\cC$ and $\cCt$ are decided by their
$\beta$-rules.

We use the local morphism\cite{abbott2003category} (the same one for encoding the Sealing
judgements) to translate inheritance judgement. This is mainly due to
the existence of \ruleref{Inh} rule. 
% Here we only typeset the overriding
% and inheritance rule, but all of the translation for inh judgment are
% induced by the $\beta$-rule of \ruleref{Inh}.

We omit all the equational rules ($\beta,\eta$ and substitution) here.
However, for coherence, when we mutual recursively define
$\denotesT{\goodSig{\Gamma}{\_}{n}}$, ${\Sigr{\Gamma}{n}} $, and
$\denotesT{\cC}$ above, we actually have to prove the substitution laws
$(\cC\sigma)[\tau]\equiv \cC(\sigma[\tau])$ and $\Sigr{\Delta}{n}[\tau]
\equiv \Sigr{\Gamma}{n}$ together. 
% The details (of all the equational rules) can be found in the appendices written in (pseudo-)Agda form. We encourage the reader to read that (pseudo-)Agda-style proof because it is more friendly to type theorists, easy to do type checking and
% validate the proof and it is clearer on the mathematical object each
% translation uses.

>>>>

The implementation described in \cref{sec:coqimpl} has hinted the fact that
linkage can be translated into dependent tuple type.
The described compilation strategy can be considered as
a \textit{type preserving syntactic translation}, a function mapping from our syntactic
model(typed Abstract Syntax Tree(AST)) to a ``subpart''\footnote{Not exactly subpart---the only
difference lies in "Wrec". In our metatheory, it is constructed
using linkage; in ``subpart'', it will be using $\Sigma$-types (translation
of the linkage).} of our syntactic model(typed AST) \textit{without
linkages}.

@@

The main idea of this syntactic translation is that, we make sure
$\goodSig{\Gamma}{\nu^+ \_ }{n+1}$ and $\goodSig{\Gamma}{\mu^+ \_}{n+1}$
wrapped into a dependent pair. These two definitions decide the
syntactic translation for $\goodSig{\Gamma}{\_}{n}$ and $\cL$. The
syntactic translation for $\cP$ are decided by their
$\beta$-rules.

% We use the local morphism\cite{abbott2003category} (the same one for encoding the Sealing
% judgements) to translate inheritance judgement. This is mainly due to
% the existence of \ruleref{Inh} rule. 
% Here we only typeset the overriding
% and inheritance rule, but all of the translation for inh judgment are
% induced by the $\beta$-rule of \ruleref{Inh}.

We omit all the equational rules ($\beta,\eta$ and substitution) here.
However, for coherence, when we mutual recursively define
$\denotesT{\goodSig{\Gamma}{\_}{n}}$, ${\Sigr{\Gamma}{n}} $, and
$\denotesT{\cP}$ above, we actually have to prove the substitution laws
$\sub{(\TyTkg{\lsig})}{\gm}\equiv \TyTkg{\sub{\lsig}{\gm}}$ and $\sub{(\Sigr{\Delta}{n})}{\gm}
\equiv \Sigr{\Gamma}{n}$ together. 
% The details (of all the equational rules) can be found in the appendices written in (pseudo-)Agda form. We encourage the reader to read that (pseudo-)Agda-style proof because it is more friendly to type theorists, easy to do type checking and
% validate the proof and it is clearer on the mathematical object each
% translation uses.
